---
title: "Progress Report"
author: "Margaret Shepherd, Anina Peersen"
date: 'December 5, 2022'
output: 
  bookdown::html_document2:
    split_by: none
    toc: yes
    toc_depth: 3
    toc_float:
      toc_collapsed: true
    number_sections: false
---

<style>
body {
text-align: justify}
</style> 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

Fires = read.csv("FW_Veg_Rem_Combined.csv")
library(ggplot2)
library(tidyverse)
library(maps)
library(ggmap)
library(ggthemes)
library(usdata)
library(lubridate)
library(stringr)
library(usmap)
library(RColorBrewer)
```

```{r, echo = FALSE}
FiresClean <- Fires %>%
  mutate(discovery_month = fct_relevel(discovery_month, c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')), 
         disc_pre_month = fct_relevel(disc_pre_month, c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')), 
         days_to_containment =  str_remove(putout_time, "\\ .*")) %>%
  mutate(fire_size_class = factor(fire_size_class), 
         stat_cause_descr = factor(stat_cause_descr), 
         state = factor(state), 
         vegCat = case_when(Vegetation %in% c(1, 2, 3, 4, 5, 6, 7, 16, 17, 18, 19, 20, 21, 22) ~ 'forest', 
                            Vegetation %in% c(24, 25, 26, 27) ~ 'agricultural', 
                            Vegetation == 28 ~ 'urban', 
                            Vegetation == 23 ~ 'water', 
                            Vegetation == 13 ~ 'tundra', 
                            Vegetation %in% c(14, 15) ~ 'desert', 
                            Vegetation %in% c(9, 10) ~ 'grassland',
                            Vegetation %in% c(11, 12) ~ 'shrubland', 
                            Vegetation == 8 ~ 'savanna', 
                            Vegetation == 0 ~ 'unknown'), 
         forestType = case_when(Vegetation %in% c(2, 5, 7, 17, 20, 22) ~ 'deciduous', 
                                Vegetation %in% c(1, 3, 4, 6, 16, 18, 19, 21) ~ 'evergreen'), 
         forestLeafType = case_when(Vegetation %in% c(1, 2, 3, 5, 16, 17, 18, 20) ~ 'broadleaf', 
                                    Vegetation %in% c(4, 6, 7, 19, 21, 22) ~ 'needleleaf'), 
         causeCat = case_when(stat_cause_descr %in% c('Arson') ~ 'Intentional Human Action', 
                              stat_cause_descr %in% c('Campfire', 'Children', 'Equipment Use', 'Fireworks', 'Smoking') ~ 'Unintentional Human Action', 
                              stat_cause_descr %in% c('Powerline', 'Railroad', 'Structure') ~ 'Infrastructure', 
                              stat_cause_descr %in% c('Lightning', 'Debris Burning') ~ 'Natural Causes', 
                              stat_cause_descr %in% c('Miscellaneous', 'Missing/Undefined') ~ 'Other'))

FiresClean$disc_clean_date <- as_date(FiresClean$disc_clean_date, format = '%m/%d/%Y')
FiresClean$cont_clean_date <- as_date(FiresClean$cont_clean_date, format = '%m/%d/%Y')
FiresClean$disc_date_pre <- as_date(FiresClean$disc_date_pre, format = '%m/%d/%Y')
FiresClean$disc_date_final <- as_date(FiresClean$disc_date_final, format = '%m/%d/%Y %h:%M')
FiresClean$cont_date_final <- as_date(FiresClean$cont_date_final, format = '%m/%d/%Y %h:%M')
```


```{r, echo = FALSE, include = FALSE}
continentalUS <- get_stamenmap(
  bbox = c(left = -126.47, bottom = 24.21, right = -65.74, top = 49.72),
  maptype = "terrain",
  zoom = 4)

Alaska <- get_stamenmap(
  bbox = c(left = -189.23, bottom = 50.4, right = -129.38, top = 71.52),
  maptype = "terrain",
  zoom = 5)

Hawaii <- get_stamenmap(
  bbox = c(left = -160, bottom = 18.15, right = -154, top = 22.6),
  maptype = "terrain",
  zoom = 7)

PuertoRico <- get_stamenmap(
  bbox = c(left = -67.7, bottom = 17.17, right = -65, top = 19.14),
  maptype = "terrain",
  zoom = 8)
```

### Introduction

Large fires are grouped in certain vegetation types.

```{r, echo = FALSE, fig.align='center'}
FiresCleanVeg <- FiresClean %>% 
  group_by(vegCat) %>% 
  summarize(num = n())

ggplot(FiresClean, aes(x = vegCat, fill = fire_size_class)) +
  geom_bar(position = 'stack') +
  labs(x = 'Surrounding Vegetation',
       y = 'Number of Fires',
       title = 'Number of Fires by Surrounding Vegetation and Fire Size Classification') + 
  theme_minimal()

ggplot(FiresClean, aes(x = vegCat, fill = fire_size_class)) +
  geom_bar(position = 'fill') +
  labs(x = 'Surrounding Vegetation',
       y = 'Number of Fires',
       title = 'Proportion of Fires Size Classification by Surrounding Vegetation') + 
  theme_minimal()
```

The numerical bar chart shows the number of fires in the dataset, grouped by the fire size class (B through G, smallest to largest) and the type of surrounding vegetation. The proportional bar chart shows the proportion of fires with the same grouping. Certain vegetations clearly have more fires than others (shrubland, forest, and desert dominate over grassland and unknown vegetation), but those are not necessarily the most dangerous vegetations. By looking at the proportional bar chart, we can see that grassland has the highest proportion of large fires (defined for the purposes of this study as class E, F, and G fires), followed by unknown vegetation desert. 



### Spatial Analysis

#### Maps

```{r, echo = FALSE, include = FALSE}
FiresClean_class <- FiresClean %>% 
  mutate(fire_size_class = fct_recode(fire_size_class, "Class B: 1/4 acre to 10 acres" = "B", 
                            "Class C: 10 acres to 100 acres" = "C",
                            "Class D: 100 acres to 300 acres" = "D",
                            "Class E: 300 acres to 1000 acres" = "E",
                            "Class F: 1000 acres to 5000 acres" = "F",
                            "Class G: 5000+ acres" = "G"))
```

Fires are concentrated in the Southern US, but high magnitude fires and large fires are more common in the Western US and Alaska.

```{r, echo = FALSE, fig.align='center'}
# maps of fire locations with fire size class and magnitude
ggmap(continentalUS) + 
  geom_point(
    data = FiresClean_class,
    aes(x = longitude, y = latitude, color = fire_mag, size = fire_size_class),
    alpha = .2
  ) +
  scale_size_manual(values = c("Class B: 1/4 acre to 10 acres" = 0.1, 
                               "Class C: 10 acres to 100 acres" = 0.4, 
                               "Class D: 100 acres to 300 acres" = 0.7, 
                               "Class E: 300 acres to 1000 acres" = 1, 
                               "Class F: 1000 acres to 5000 acres" = 1.3, 
                               "Class G: 5000+ acres" = 1.6)) +
  theme_map() +
  theme(legend.position = "right") +
  scale_color_gradient(low="yellow", high="red")

ggmap(Alaska) +
  geom_point(
    data = FiresClean_class,
    aes(x = longitude, y = latitude, color = fire_mag, size = fire_size_class),
    alpha = .5
  ) +
  scale_size_manual(values = c("Class B: 1/4 acre to 10 acres" = 0.1, 
                               "Class C: 10 acres to 100 acres" = 0.4, 
                               "Class D: 100 acres to 300 acres" = 0.7, 
                               "Class E: 300 acres to 1000 acres" = 1, 
                               "Class F: 1000 acres to 5000 acres" = 1.3, 
                               "Class G: 5000+ acres" = 1.6)) +
  theme_map() +
  theme(legend.position = "right") +
  scale_color_gradient(low="yellow", high="red")

ggmap(Hawaii) +
  geom_point(
    data = FiresClean_class,
    aes(x = longitude, y = latitude, color = fire_mag, size = fire_size_class),
    alpha = .5
  ) +
  scale_size_manual(values = c("Class B: 1/4 acre to 10 acres" = 0.1, 
                               "Class C: 10 acres to 100 acres" = 0.4, 
                               "Class D: 100 acres to 300 acres" = 0.7, 
                               "Class E: 300 acres to 1000 acres" = 1, 
                               "Class F: 1000 acres to 5000 acres" = 1.3, 
                               "Class G: 5000+ acres" = 1.6)) +
  theme_map() +
  theme(legend.position = "right") +
  scale_color_gradient(low="yellow", high="red")

ggmap(PuertoRico) +
  geom_point(
    data = FiresClean_class,
    aes(x = longitude, y = latitude, color = fire_mag, size = fire_size_class),
    alpha = .5
  ) +
  scale_size_manual(values = c("Class B: 1/4 acre to 10 acres" = 0.1, 
                               "Class C: 10 acres to 100 acres" = 0.4, 
                               "Class D: 100 acres to 300 acres" = 0.7, 
                               "Class E: 300 acres to 1000 acres" = 1, 
                               "Class F: 1000 acres to 5000 acres" = 1.3, 
                               "Class G: 5000+ acres" = 1.6)) +
  theme_map() +
  theme(legend.position = "right") +
  scale_color_gradient(low="yellow", high="red")
```


The locations of the dots indicate a fire in the dataset, the color indicates the fire magnitude (redder color corresponds to higher magnitude), and the size indicates the fire size class (B - G corresponds to smaller - larger). The dots are most highly concentrated in the South (with a path carved out by the Mississippi River), but light colored dots are scattered throughout the Southwest and Pacific Northwest, with a populations also scattered through Texas and Florida. Alaska also has a high concentration of high magnitude fires. When combating the most dangerous and destructive fires, efforts should be focused on Alaska, the Southwest, and the Pacific Northwest.
 

Fires are concentrated in the Southern US, but large fires are more common in the Western US and Alaska.

```{r, echo = FALSE, fig.align='center'}
# maps of fire locations and fire class size
ggmap(continentalUS) + 
  geom_point(
    data = FiresClean_class,
    aes(x = longitude, y = latitude, color = fire_size_class),
    alpha = .4,
    size = 0.2
  ) +
  theme_map() +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 0.8))) +
  theme(legend.position = "right")

ggmap(Alaska) +
  geom_point(
    data = FiresClean_class,
    aes(x = longitude, y = latitude, color = fire_size_class),
    alpha = .8,
    size = 0.7
  ) +
  theme_map() +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 0.8))) +
  theme(legend.position = "right")

ggmap(Hawaii) +
  geom_point(
    data = FiresClean_class,
    aes(x = longitude, y = latitude, color = fire_size_class),
    alpha = .8,
    size = 1
  ) +
  theme_map() +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 0.8))) +
  theme(legend.position = "right")

ggmap(PuertoRico) +
  geom_point(
    data = FiresClean_class,
    aes(x = longitude, y = latitude, color = fire_size_class),
    alpha = .8,
    size = 1
  ) +
  theme_map() +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 0.8))) +
  theme(legend.position = "right")
```

The pattern shown in this set of maps echoes the message shown in the above maps. Fires overall are mostly located in the South, but large fires (displayed by the purple and blue dots) are most common in the Southwest, Pacific Northwest, Texas, Florida, and Alaska.


On a state-by-state basis, most wildfires in the US are started by natural causes.

```{r, echo = FALSE, fig.align='center'}
# MOST COMMON CAUSE OF FIRE BY STATE
FiresCleanState <- FiresClean %>% 
  group_by(state, causeCat) %>% 
  summarize(num_by_cause = n())

FiresCleanGroup <- FiresCleanState %>% 
  group_by(state) %>% 
  filter(num_by_cause == max(num_by_cause)) # need to deal with MA, DE

FiresCleanGroup <- FiresCleanGroup %>%
  ungroup() %>% 
  filter(!state %in% c('DE', 'MA')) %>% # takes out DE and MA
  add_row(state = "DE", # re-adds DE
          causeCat = "Intentional Human Action, Unintentional Human Action, Other", 
          num_by_cause = 3) %>%
  add_row(state = "MA", # re-adds MA
          causeCat = "Unintentional Human Action, Other", 
          num_by_cause = 24) 
  
plot_usmap(data = FiresCleanGroup, values = "causeCat", color = "black") +
  scale_fill_manual(values = c("#1b6c6b","green", "#1ccbd1", "#bcdfc6", "#e8f7eb"),
                    # https://huemint.com/gradient-5/ #00a7a0
                    breaks = c("Natural Causes", 
                               "Other", 
                               "Intentional Human Action", 
                               "Unintentional Human Action, Other",
                               "Intentional Human Action, Unintentional Human Action, Other"),
                    labels = c("Natural Causes: lightning, debris burning", 
                               "Other: miscellaneous, missing/undefined", 
                               "Intentional Human Action: arson", 
                               "Tie: Unintentional Human Action, Other (MA)",
                               "Tie: Intentional Human Action, Unintentional Human Action, Other (DE)"),
                    name = "Cause of Wildfire") +
  theme(legend.position = "right") +
  labs(title = "Most Common Cause of Wildfire by State (1992-2015)", 
       caption = "Data from federal, state, and local fire organizations in the US. \n Graphic made by Margaret Shepherd. \n Unintentional Human Action consists of campfires, children, equipment use, \n fireworks, and smoking.")
```

For most states in the US, the most common cause of wildfires is natural causes. A smaller group of states has “other” as the leading cause, and four states list arson as the most common cause of wildfires. It seems that in most places, we can only control our preparation for and response to wildfires, not whether or not they occur in the first place, but this graph may pinpoint states where wildfire records are unclear or underutilized, and states where the most effective firefighting strategy is tied to crime reduction.

#### Bar Plots

The states with the largest fires are concentrated in the western half of the US.

```{r, echo = FALSE, fig.align='center'}
# large fires
LargeFires_state <- FiresClean %>%
  filter(fire_size_class == "E" | fire_size_class == "F" | fire_size_class == "G") %>% 
  group_by(state) %>% 
  summarize(number = n()) %>% 
  arrange(desc(number)) %>% 
  head(10)

ggplot(LargeFires_state, aes(x = reorder(state, desc(number)), y = number)) +
  geom_col() +
  labs(x = 'State',
       y = 'Number of Large (300+ acres) Wildfires',
       title = 'Top 10 US States with the Most Large Wildfires from 1992-2015')
```

The states with the largest fires (fires at or above 300 acres) are mostly in the western half of the US, with outliers at the top of the graph (Alaska) and bottom of the graph (Florida). This is only a generalization, since Washington, Utah, Colorado, and Wyoming do not make the cut. 

### Temporal Analysis

As shown in the plot below showing all fires in our dataset, fires are generally cyclical with spikes every 4-6 years. This cycle generally corresponds to natural cycles in precipitation levels, temperature, and global cycles such as El Niño phases. We are looking into ways to visualize fire cycles alongside these other cycles. 

```{r, echo = FALSE, fig.align='center'}
ggplot(FiresClean, aes(x = disc_clean_date)) + 
  geom_density(fill = 'orangered', color = NA, alpha = .7) + 
  labs(x = "Date of Fire Discovery", y = "Comparative Density", title = "Density of Fires, 1992-2015") + 
  theme_minimal() 
```


#### Weather Immediately Before a Fire

```{r, echo = FALSE, include = FALSE}
Fires_temp <- FiresClean %>% 
  filter(weather_file != 'File Not Found') %>% 
  filter(Temp_pre_30 != 0) %>% 
  filter(Temp_pre_15 != 0) %>%
  filter(Temp_pre_7 != 0) %>%
  filter(Temp_cont != 0) 
  # filter(Temp_pre_30 != 0 | Temp_pre_15 != 0 | Temp_pre_7 != 0 | Temp_cont != 0) # keeps row if all four temps are nonzero

Fires_wind <- FiresClean %>% 
  filter(weather_file != 'File Not Found') %>%
  filter(Wind_pre_30 != 0) %>% 
  filter(Wind_pre_15 != 0) %>%
  filter(Wind_pre_7 != 0) %>%
  filter(Wind_cont != 0)
  
Fires_hum <- FiresClean %>% 
  filter(weather_file != 'File Not Found') %>%
  filter(Hum_pre_30 != 0) %>% 
  filter(Hum_pre_15 != 0) %>%
  filter(Hum_pre_7 != 0) %>%
  filter(Hum_cont != 0)

Fires_prec <- FiresClean %>% 
  filter(weather_file != 'File Not Found') %>%
  filter(Prec_pre_30 != 0) %>% 
  filter(Prec_pre_15 != 0) %>%
  filter(Prec_pre_7 != 0) %>%
  filter(Prec_cont != 0)

(avgtemp30 = mean(Fires_temp$Temp_pre_30))
(avgtemp15 = mean(Fires_temp$Temp_pre_15))
(avgtemp7 = mean(Fires_temp$Temp_pre_7))
(avgtempcont = mean(Fires_temp$Temp_cont))

(avgwind30 = mean(Fires_wind$Wind_pre_30))
(avgwind15 = mean(Fires_wind$Wind_pre_15))
(avgwind7 = mean(Fires_wind$Wind_pre_7))
(avgwindcont = mean(Fires_wind$Wind_cont))

(avghum30 = mean(Fires_hum$Hum_pre_30))
(avghum15 = mean(Fires_hum$Hum_pre_15))
(avghum7 = mean(Fires_hum$Hum_pre_7))
(avghumcont = mean(Fires_hum$Hum_cont))

(avgprec30 = mean(Fires_prec$Prec_pre_30))
(avgprec15 = mean(Fires_prec$Prec_pre_15))
(avgprec7 = mean(Fires_prec$Prec_pre_7))
(avgpreccont = mean(Fires_prec$Prec_cont))

days_before <- c(30, 15, 7, 0)
temp <- c(avgtemp30, avgtemp15, avgtemp7, avgtempcont)
wind <- c(avgwind30, avgwind15, avgwind7, avgwindcont)
hum <- c(avghum30, avghum15, avghum7, avghumcont)
prec <- c(avgprec30, avgprec15, avgprec7, avgpreccont)
avg_weather <- data.frame(days_before, temp, wind, hum, prec)
```

Decreasing precipitation in the 30 days preceding a wildfire is the most predictive weather phenomenon.

```{r, echo = FALSE, fig.align='center'}
ggplot(data = avg_weather, aes(x = days_before)) +
  geom_point(aes(y = temp/max(temp), color = "Temperature")) +
  geom_point(aes(y = wind/max(wind), color = "Wind")) +
  geom_point(aes(y = hum/max(hum), color = "Humidity")) +
  geom_point(aes(y = prec/max(prec), color = "Precipitation")) +
  geom_line(aes(y = temp/max(temp), color = "Temperature")) +
  geom_line(aes(y = wind/max(wind), color = "Wind")) +
  geom_line(aes(y = hum/max(hum), color = "Humidity")) +
  geom_line(aes(y = prec/max(prec), color = "Precipitation")) + 
  xlim(30, 0) +
  labs(x = 'Number of Days Before a Wildfire',
       y = 'Normalized Average Weather at Wildfire Location',
       title = 'Average Temperature, Wind Speed, Humidity, and Precipitation Before a Wildfire \n Wildfires in the US, 1992-2015',
       color = "Weather") +
  scale_color_manual(values=c("purple", "blue", "red", "orange")) + 
  theme_minimal()
```

This plot was made by averaging the temperature, wind speed, humidity, and precipitation at certain intervals leading up to a wildfire and normalizing the values by the largest value in each respective dataset, in order to better visualize the average change in each. Precipitation by far shows the largest change leading up a fire, making it the most potent predictor. Humidity and wind speed also decrease leading up to a fire while temperature increases. An observed combination of these factors can help increase awareness of and preparation for high fire risk.

#### Global Weather Patterns

Two global weather patters with effects on fire likelihood are El Niño and La Niña. They are two sides of the same coin: El Niño is the warm counterpart and La Niña the cold. Overall, El Niño starts with above average ocean surface temperatures in the central and eastern Pacific Ocean. Globally, El Niño causes the low-level trade winds that blow east to west along the equator weaken or even blow the opposite direction. Understandably, this has massive implications for global weather trends. For instance, there is reduced rainfall over Southeast and South Asia and increased rainfall - and subsequently increased tropical storm formation - in the greater Pacific. In the US, El Niño increases precipitation over the Gulf Coast and decreases precipitation in Hawaii, the Ohio Valley, the Pacific Northwest, and the Rocky Mountains. On the other hand, La Niña is the opposite: it starts with below average ocean surface temperatures in the central and eastern Pacific which leads to heavy rains over Southeast Asia and Southern Africa, while Northwest Africa will be drier than usual. In the US, La Niña increases precipitation in the Northern Midwest, the Northern Rocky Mountains, the Pacific Northwest, and Northern California. Simultaneously, precipitation decreases across the entire southern US, from California to Florida. Both of these weather patterns happen approximately every 2-7 years and their effects can be anywhere from 6 months to 2 years. 

Given this basic understanding of El Niño and La Niña, it is clear that they affect fire trends, as they both affect the leading predictor of fire, precipitation. Below, you can see our spotlight on natural fires in Idaho and Texas, as both states have some of the most fires in the US and have opposite effects of El Niño and La Niña. We limited this graph to natural fires as global meteorological trends will primarily impact on natural fires. 


```{r, echo = FALSE, fig.align = 'center'}
idatexnatural <- FiresClean %>%
  filter(state %in% c("ID", "TX")) %>%
  filter(causeCat == "Natural Causes")

LaNina2010 <- data.frame(xstart = as.POSIXct('2010-07-01'), xend = as.POSIXct('2012-07-01'))
LaNina2008 <- data.frame(xstart = as.POSIXct('2008-07-01'), xend = as.POSIXct('2009-07-01'))
LaNina2005 <- data.frame(xstart = as.POSIXct('2005-07-01'), xend = as.POSIXct('2006-07-01'))
LaNina1998 <- data.frame(xstart = as.POSIXct('1998-07-01'), xend = as.POSIXct('2001-07-01'))
LaNina1995 <- data.frame(xstart = as.POSIXct('1995-07-01'), xend = as.POSIXct('1996-07-01'))

ElNino1997 <- data.frame(xstart = as.POSIXct('1997-06-01'), xend = as.POSIXct('1998-06-01'))
ElNino2002 <- data.frame(xstart = as.POSIXct('2002-06-01'), xend = as.POSIXct('2003-06-01'))
ElNino2004 <- data.frame(xstart = as.POSIXct('2004-06-01'), xend = as.POSIXct('2005-06-01'))
ElNino2006 <- data.frame(xstart = as.POSIXct('2006-06-01'), xend = as.POSIXct('2007-06-01'))
ElNino2009 <- data.frame(xstart = as.POSIXct('2009-06-01'), xend = as.POSIXct('2010-06-01'))
ElNino2014 <- data.frame(xstart = as.POSIXct('2014-05-01'), xend = as.POSIXct('2015-12-31'))

ggplot() + 
  geom_rect(data = LaNina2010, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .75, fill = 'blue') + 
  geom_rect(data = LaNina2008, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .4, fill = 'blue') +
  geom_rect(data = LaNina2005, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .4, fill = 'blue') + 
  geom_rect(data = LaNina1998, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .4, fill = 'blue') + 
  geom_rect(data = LaNina1995, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .4, fill = 'blue') + 
  geom_rect(data = ElNino1997, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .75, fill = 'purple') + 
  geom_rect(data = ElNino2002, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .4, fill = 'purple') + 
  geom_rect(data = ElNino2004, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .4, fill = 'purple') +
  geom_rect(data = ElNino2006, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .4, fill = 'purple') +
  geom_rect(data = ElNino2009, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .4, fill = 'purple') +
  geom_rect(data = ElNino2014, aes(xmin = xstart, xmax = xend, ymin = 0, ymax = Inf), alpha = .75, fill = 'purple') +
  geom_density(data = idatexnatural, aes(x = as.POSIXct(disc_clean_date), fill = state), alpha = .65, color = 'white') + 
  scale_fill_manual(name = NULL, values = c('orangered', 'orange'), labels = c("Idaho", "Texas")) +
  xlim(as.POSIXct('1992-01-01'), as.POSIXct('2015-12-31')) + 
  labs(x = 'Date of Fire Discovery', y = 'Comparative Density', title = 'Natural Fires in Idaho and Texas with La Niña and El Niño Events, 1992-2015', subtitle = 'La Niña events appear in blue; El Niño events appear in purple. The La Niña event starting\n2010 was particularly intense and is therefore represented with a darker blue; the same for\nEl Niño events starting in 1997 and 2014.') + 
  theme_minimal() + 
  theme(legend.position = "bottom") 
```

As shown on the graph, the three low-level fire density spikes in Idaho correspond to La Niña events. However, interesting the 2010-2012 event did not correspond to a larger spike in fires, despite being more intense. This result is also interesting because La Niña causes increased precipitation in Idaho, not decreased. We would expect to see decreased precipitation with spikes in fires. However this could be because the increased rainfall would allow undergrowth to thrive, which would then heighten fire risk when the rains stopped and the plants were more dried out. 

In Texas, the three large spikes in the late 2000s and early 2010s also all correspond with La Niña events. Even the small spike in the late 1990s corresponds with a La Niña event. However in Texas, the spikes generally occur during the event, as opposed to after the event in Idaho. This is most likely because La Niña causes decreased precipitation in Texas during the event, and since Texas is already a relatively dry state, the decrease in precipitation increases fire risk faster. 

Finally, the strong El Niño event starting in 2014 corresponds to sharp drops in natural fires in both states. In Texas, El Niño causes increased precipitation, which would understandably correspond with a drop in fires and such a sharp drop was mostly likely due to the increased intensity and length of the event. However in Idaho, El Niñ corresponds to less precipitation. The drop in fires corresponds to a predictible drop after a spike, however the drop seems to continue further than expected given lower precipitation. However, since this event spans the end of our study, it is difficult to make any definitive statements. 


#### By Fire Size Class

As shown in the faceted density plot below, fires are generally cyclical. Spikes occur approximately every 5 years, regardless of size classification.

```{r, echo = FALSE, fig.align='center'}
ggplot(FiresClean, aes(x = disc_clean_date, fill = fire_size_class)) + 
  geom_density(alpha = .6, color = NA) + 
  facet_wrap(~ fire_size_class) + 
  scale_fill_manual(values = c('orange', 'darkorange', 'darkorange2', 'orangered', 'red', 'brown')) + 
  labs(x = 'Date of Fire Discovery', y = 'Comparative Density', title = "Fire Density by Size Classification, 1992-2015") +
  theme_minimal() + 
  theme(legend.position = 'none')
```

Furthermore, when aggregated by day, small fires spike in the spring and large fires spike in the late summer. Intermediate fires spike in both the spring and late summer.

```{r, echo = FALSE, fig.align='center'}
ggplot(FiresClean, aes(x = yday(disc_clean_date), fill = fire_size_class)) + 
  geom_density(alpha = .6, color = NA) + 
  facet_wrap(~ fire_size_class) + 
  scale_fill_manual(values = c('orange', 'darkorange', 'darkorange2', 'orangered', 'red', 'brown')) + 
  scale_x_continuous(breaks = c(1, 60, 121, 182, 244, 305), labels = c('Jan', 'Mar', 'May', 'Jul', 'Sept', 'Nov')) +
  labs(x = 'Date of Fire Discovery', y = 'Comparative Density', title = "Fire Density by Size Classification, Aggregated by Day") + 
  theme_minimal() + 
  theme(legend.position = 'none')
```

The distribution of fires over our entire study shows that fires are cyclical approximately every 5 years. This would presumably correspond to yearly patterns in drought and heat as well was natural regrowth and revival. This information is useful in predicting the fire risk for a year based on fires from the previous year. Furthermore, when aggregated across the data set to the day of the year, we see that smaller fires spike in the spring, larger fires spike in the late summer and early fall, and intermediate fires have dual spikes in both. This is also useful for determining fire risk within a year. 


#### By Cause of Fire

Now to look at the temporal distribution of fire by cause, we made the below density plot of cause over the entire study range. The most important take-away is again the cyclical nature of the fires, most noticeable in fires caused by either natural causes or unintentional human actions. Infrastructure- and intentionally-caused fires have much less noticeable patterns. Infrastructure fires remained fairly constant except the large drop from 2000-2005. After this drop, fire density increased fairly steadily, reaching pre-2000 levels around 2011, but subsequently dropped even lower than previously. Fires caused by intentional human action spiked around 2000 and had a second smaller spike around 2008. We are currently unsure why these spikes may have occurred, but are investigating further. Fires with other causes had a large spike around 2006 and a second spike around 2011. We are also unsure of causes for this spike, but will most likely not be able to find possible causes as the causes of these fires were largely not included in our dataset. 

```{r, echo = FALSE, fig.align='center'}
ggplot(FiresClean, aes(x = disc_clean_date, fill = causeCat)) + 
  geom_density(alpha = .6, color = NA) + 
  facet_wrap(~ causeCat) + 
  labs(x = 'Date of Fire Discovery', y = 'Comparative Density', fill = 'Fire Cause', title = "Fire Density by Cause, 1992-2015") +
  theme_minimal() + 
  theme(legend.position = 'none') 
```


When the fire density is aggregated across our time frame by day, we see infrastructure- and intentionally-set fires, as well as those with other causes, spike in the spring. Fires started by natural causes have two spikes: in the spring and the late summer. This dual spike most likely corresponds to the different spikes for fire size as mentioned previously. Unintentional human action caused fires spike in the spring and again in the summer, with the latter spike most likely caused by fireworks on the 4th of July. 

```{r, echo = FALSE, fig.align='center'}
ggplot(FiresClean, aes(x = yday(disc_clean_date), fill = causeCat)) +
  geom_density(alpha = .6, color = NA) + 
  labs(x = 'Date of Fire Discovery', y = 'Comparative Density', title = 'Fire Density by Cause, Aggregated by Day') +
  facet_wrap(~ causeCat) + 
  scale_x_continuous(breaks = c(1, 60, 121, 182, 244, 305), labels = c('Jan', 'Mar', 'May', 'Jul', 'Sept', 'Nov')) +
  theme_minimal() + 
  theme(legend.position = 'none')
```

These density plots in and of themselves lend important insights into fire prevalence. However, the most important information for fire-prediction comes from combining the information here with that shown in the density by cause plots. 




